---
title: 'Analysis: Road Casualties - Camden'
author: "[Avision Ho](https://github.com/avisionh)"
date: "`r Sys.Date()`"
output: 
  html_document:
    number_sections: false
    theme: journal
    highlight: haddock
    code_folding: hide
    fig_caption: true
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(results = "asis", warning = FALSE, message = FALSE)

# Load packages
library(dplyr)
library(ggplot2)
library(pander)

# Turn off scientific notation
options(scipen = 999, expressions = 10000)

# Set theme for plots
plot_theme <- theme(plot.title = element_text(face = "bold", hjust = 0.5),
                    plot.subtitle = element_text(face = "bold", hjust = 0.5),
                    panel.background = element_blank(),
                    axis.line = element_line(color = "black"))

```

# Section 1: Executive Summary
In the exploratory analysis to follow, we found:

1. Of those with a road casualty, if they were a man, then they were more likely to have been travelling via a *Powered 2-wheeler* whereas a woman would have more likely been a `Pedestrian`.
1. Continuing with this analysis of sex, the general age-(interquartile-)range for men was higher for fatal incidences of road casualty than women, though men did display a lower average age for *Fatal* and *Serious* casualties. 
1. From the *Fatal* and *Serious* incidences of casaulty, the *Driver/Rider* and *Pedestrians* were more likely to suffer these than a *Passenger* would. This could be for a variety of reasons, and also corroborates point (1.) earlier.
1. In terms of the time period and which days casualties are most likely to take place, with the exception of Friday and the weekend, casualties typically happen during the rush hour period after work.
1. Casting attention to the location of these fatal or serious casualties, most of them take place in the bottom centre of the Camden Borough of London. These aligns with the main throughfare of British Museum, British Library and Great Ormond Street hospital.
1. Looking at the impact of the weather, it does not seem to have impacted the casualties too much as most casualties took place in *Fine* weather where there was *Daylight*.

Moving forward, can apply data science techniques such as **K-Nearest Neighbour** or **Logistic Regression** to understand the groups of people more likely to be at risk of casualties and to appreciate the factors that can increase the chances of casualties occurring.

# Section 2: Background

## Introduction
The purpose of this project is to explore the road casualty data publicly available for the London Borough of Camden.

## Motivation
By exploring the data, we hope to identify and understand the following:

1. **Who** are most involved in road casualties

1. **What** types of casualties are most frequent, including their severity

1. **When** these casualties often occur 

1. **Where** road casualties take place

1. **How** these casualties take place, including weather conditions

The idea behind this exploratory analysis is to provide the platform for a algorithm to be applied to better understand who, what, when, where and how these casualties take place so more effective campaigns can be delivered by the local council to reduce casualties as well as identifying *hotspots* for where casualties most likely happen so that work such as infrastructure builds to improve the roads, can be done by the council.

## Data
The data for road casualties comes from the data.gov.uk website [here](https://data.gov.uk/dataset/f36669c1-fe1f-4cbc-803c-db49675247c3/road-collision-casualties-in-camden).

The data for mapping the London Borough of Camden comes from Camden council's open data portal [here](https://opendata.camden.gov.uk/Maps/Camden-Ward-Boundary/yqyi-6agf).

This is open data as signalled by it being publicly available and having the open data license. 

### Data Quality Assessment
Owing to the nature of open data, then the expectation is that it will be of relatively poor quality and fairly messy. This is because oftentimes not as much effort is put into improving the standard, quality, accuracy and neatness of data available in the public domain.

However, given that this data comes from a government body (technically MCHLG), then there is a minimum expectation that what's being publicly provided, whilst not to the same standard as a statistical publication, is of a sufficient quality to enable relatively better analysis compared to other open data sources.

Additionally, the data is remarkably neat and adheres to tidy data principles, which suggests some processing was done in the background to format the data into something easy to consume for the public. This also suggests that some time was also spent in ensuring the data is as accurate as is feasibly possible.

With more time, one would have liked to model the unknowns entries to see how many of these we have. They were not encoded consistently in the data. Moreover, one would have liked to explore population data for the borough alongside other similar boroughs to see how well their performance compares on road casualties.


```{r (HIDE) Data Preparation, echo = FALSE, include = FALSE}
source("scripts/0_data_wrangle.R")
source("scripts/3_4_analysis_where.R")
```

# Section 3: Analysis

## Who are most involved in road casualties?
We will begin by exploring the charateristics of those people who were involved in road casualties.

```{r (SHOW) Who are involved, echo = FALSE}
plot_base <- ggplot(data = data_casualty, mapping = aes(x = ModeOfTravel, fill = CasualtySex)) +
  labs(x = "Mode of Travel") +
  plot_theme +
  # add text-wrapping for x labels
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10))
```

From the bar-chart below, we see that *men* are involved in casualties relating to *Powered 2-wheeler* whereas *women* are more involved in *Pedestrian* casualties.

```{r (SHOW) Who are involved - mode of travel, echo = FALSE}
plot_base +
  geom_bar(position = position_dodge()) +
  labs(title = "Bar-Chart: Count of Road Casualties", 
       subtitle = "by Mode of Travel and Sex",
       y = "Count of casualties")
```

Let's drill-down further as we have age related information in the dataset.

What we see in the box-plot is that for *fatal* casualties, they tend to be male who are younger than their female counterparts, on average. Yet the (interquartile-) range of ages of males involved in a *fatal* road casualty is higher than their female counterparts - this is opposite to what we see in the *serious* road casualty severity. Though in both cases, the average age for males is lower than the females.

```{r (SHOW) Who are involved - age, echo = FALSE}
ggplot(data = data_casualty, mapping = aes(x = CasualtySeverity, y = CasualtyAge, fill = CasualtySex)) +
  geom_boxplot() +
  labs(title = "Box-Plot: Severity of Casualty by Age",
       x = "Severity of Casualty",
       y = "Age of person involved in casualty") + 
  plot_theme
```

## What types of casualties are most frequent?
We will now investigate what types of casualties are most frequent so we can understand the actual end-product which impacts on casualties. For instance, if cars were the most likely cause behind the casualties, then would recommend avoiding driving in a public safety campaign.

From the table below, we see that for the *serious* and *fatal* severity casualties, the most frequent cases of these were from:

- Driver/Rider
- Pedestrian

```{r (SHOW) What types of casualties - table, echo = FALSE}
data_temp <- data_casualty %>% 
  group_by(CasualtySeverity, CasualtyClass) %>% 
  summarise(Total = n()) 

data_temp %>% 
  pander()
```

```{r (HIDE) What types of casualties - words, include = FALSE, echo = FALSE}
data_temp <- data_temp %>% 
  filter(CasualtyClass %in% c("1 Driver/Rider", "3 Pedestrian") & CasualtySeverity %in% c("1 Fatal", "2 Serious")) %>% 
  group_by(CasualtyClass) %>% 
  summarise(Total = sum(Total))

txt_driver <- data_temp %>% 
  filter(CasualtyClass == "1 Driver/Rider") %>% 
  select(Total) %>% 
  pull()
txt_pedestrian <- data_temp %>% 
  filter(CasualtyClass == "3 Pedestrian") %>% 
  select(Total) %>% 
  pull()

if (txt_driver > txt_pedestrian) {
  txt_compare <- "more"
} else if (txt_driver == txt_pedestrian) {
  txt_compare <- "same"
} else {
  txt_compare <- "less"
}
  
```

In total, there are `r toString(txt_driver)` *Driver/Rider* people who either experienced a *fatal* or *serious* casualty. This figure was `r toString(txt_pedestrian)` for *Pedestrian*s. Therefore, there were `r toString(txt_compare)` *Driver/Rider* *fatal*/*serious* casualties relative to *Pedestrians*.


## When do these casualties most frequently occur?
In this part, we will look at the times and days that these casualties take place so we can understand what days and times are most likely to yield a casualty and thereby allowing the emergency services in the council to respond approrpriately.

Looking at the plots below for each day, we learn that casulaties happen most likely around the periods:

- **Monday**: it's around rush hour going to work time.
- **Tuesday**: it's around rush hour going home time,
- **Wednesday**: it's around the same but more accidents overall compared to previous two days.
- **Thursday** again around home time and less casualties than Wednesday.
- **Friday**: around late evening as people go out to club and drink (possibly this is where expect most casualties overall).
- **Saturday and Sunday**: is calmer, especially on Saturday morning.

```{r (SHOW) When do these casualties occur - time-series, echo = FALSE}
data_temp <- data_casualty %>% 
  filter(CasualtyClass %in% c("1 Driver/Rider", "3 Pedestrian") & CasualtySeverity %in% c("1 Fatal", "2 Serious")) 

ggplot(data = data_temp, mapping = aes(x = Time, colour = Day)) +
  geom_freqpoly(binwidth = 0.1, show.legend = FALSE) +
  facet_wrap(facets = vars(Day)) +
  labs(title = "Line-Chart: Fatal Or Severe Casualties across Days and Time",
       x = "Time during the day",
       y = "Count of casualties") +
  plot_theme
```

## Where are these road casualties taking place?
In this section, will visualise where the road casualties take place in Camden so one can identify hotpots of casualties that may need thorough investigation into.

What we have below shows that down the centre of Camden, we have a higher concentration of casualties, suggesting more densely populated areas with drivers/riders and pedestrians.

```{r (SHOW) Where are road casualties occurring, echo = FALSE}
ggplot() + 
  geom_polygon(data = shape_camden_data, mapping = aes(x = long, y = lat, group = group, fill = id)) +
  geom_point(data = data_temp, mapping = aes(x = Longitude, y = Latitude)) +
  coord_fixed(1.3) +
  guides(fill = FALSE) +
  # tried to add text of location but doesn't exist
  #with(centroids, annotate(geom="text", x = long, y = lat, label=label, size=2)) +
  labs(title = "Location of Fatal or Serious Casualties") +
  theme_bare
```

## How did these casualties take place?
In this section, will explore possible factors that can cause these casualties to occur.

One suspects that `Weather` and `RoadSurface` are intrinsically related so we will not include both together in analysis to avoid double-counting.

From the plot below, the highest proportion of those suffering from **fatal** or **Serious** casualties went through *Fine* conditions during *Daylight* which corrobarates with what we found on all days except for Friday when the peak period when people would suffer from a casualty is after work during the rush hour back home.

```{r (SHOW) How did casualties occur, echo = FALSE}
data_temp <- data_temp %>% 
  count(Weather, LightConditionBand) %>% 
  mutate(Proportion = prop.table(n))

ggplot(data = data_temp, mapping = aes(x = Weather, y = Proportion, fill = LightConditionBand, label = scales::percent(Proportion))) +
  geom_col(position = "dodge") +
  geom_text(position = position_dodge(width = .9),    # move to center of bars
            vjust = -0.5,    # nudge above top of bar
            size = 3) + 
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Proportion of Fatal or Serious Casualties \nby Weather and Light Band Condition") +
  plot_theme +
  # add text-wrapping for x labels
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10))
```

# Section 4: Conclusions
Following from the exploratory analysis, we found:

1. Of those with a road casualty, if they were a man, then they were more likely to have been travelling via a *Powered 2-wheeler* whereas a woman would have more likely been a `Pedestrian`.
1. Continuing with this analysis of sex, the general age-(interquartile-)range for men was higher for fatal incidences of road casualty than women, though men did display a lower average age for *Fatal* and *Serious* casualties. 
1. From the *Fatal* and *Serious* incidences of casaulty, the *Driver/Rider* and *Pedestrians* were more likely to suffer these than a *Passenger* would. This could be for a variety of reasons, and also corroborates point (1.) earlier.
1. In terms of the time period and which days casualties are most likely to take place, with the exception of Friday and the weekend, casualties typically happen during the rush hour period after work.
1. Casting attention to the location of these fatal or serious casualties, most of them take place in the bottom centre of the Camden Borough of London. These aligns with the main throughfare of British Museum, British Library and Great Ormond Street hospital.
1. Looking at the impact of the weather, it does not seem to have impacted the casualties too much as most casualties took place in *Fine* weather where there was *Daylight*.

## Next Steps - suggestions for data science techniques
To take this analysis further, enough variables have been covered to provide a good starting point for choosing variables to include in a classification model such as k-nrearest neighbour to understand the groups of people more at risk of road casualties in the London Borough of Camden. These variables can be:

- CasualtyClass
- CasualtySex
- CasualtyAgeBand/CasualtyAge
- PedestrianLocation
- PedestrianMovement
- ModeOfTravel

Even more powerful is building a predictive model such as a logistic regression or a random forest to understand the impacts the following variables can have on road casualty:

- ModeOfTravel
- NumberofVehicles
- Day
- Time
- Road Type
- Speed Limit
- LightConditionBand
- Weather
- Weather (though likely to be highly correlated with RoadSurface, so worth only including one of these only)
- RoadSurface